---
title: "Git & GitHub for Scientists — Welcome & Setup"
subtitle: "NIEHS 3-Hour Workshop on Posit Cloud"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Welcome! {.unnumbered}

Welcome to **Git & GitHub for Scientists**, a hands-on workshop for researchers
at the National Institute of Environmental Health Sciences (NIEHS).

By the end of this workshop you will be able to:

- Track every change to your analysis code with Git
- Collaborate with colleagues through GitHub pull requests and issues
- Create branches to safely experiment without breaking working code
- Write commit messages that document your scientific reasoning
- Keep sensitive data out of version control
- Create citable code releases with DOIs via Zenodo

## Workshop Agenda

| Module | Topic | Time |
|--------|-------|------|
| 0 | Welcome & Setup (this notebook) | 10 min |
| 1 | Why Version Control for Science? | 20 min |
| 2 | Setting Up: GitHub, PAT, & Posit Cloud | 25 min |
| 3 | The Core Git Workflow | 40 min |
| — | Break | 10 min |
| 4 | Branching & Collaboration | 30 min |
| 5 | GitHub for Scientific Teams | 25 min |
| 6 | Best Practices & .gitignore | 15 min |
| — | Capstone Exercise | 15 min |


# Prerequisites Checklist

Please confirm you have the following ready before we begin:

```{r checklist, echo=FALSE, results='asis'}
items <- c(
  "A free **GitHub** account — [github.com](https://github.com)",
  "A free **Posit Cloud** account — [posit.cloud](https://posit.cloud)",
  "A **modern web browser** (Chrome, Firefox, or Edge)",
  "Your **NIEHS email** address handy"
)
cat(paste0("- [ ] ", items, "\n"))
```

> **No prior Git experience is needed.** If you have used R and RStudio before,
> you have all the background necessary.


# How to Use These Notebooks

Each notebook in this workshop is an R Markdown document. You are learning
about Git *while using R Markdown* — two powerful reproducibility tools at once.

**Two ways to work through the material:**

1. **Read along** as the instructor presents the slides
2. **Run the exercises** by typing Git commands in the **Terminal** tab
   (bottom-left pane → Terminal, NOT the Console)

## Terminal vs. Console — Critical Distinction

This is the single most common source of confusion for beginners:
 
| | **Terminal** | **Console** |
|---|---|---|
| Prompt | `$` | `>` |
| Runs | Bash / shell commands | R code |
| Used for | **Git commands** | Data analysis |
| Access | Bottom-left → Terminal tab | Bottom-left → Console tab |

**If you type `git status` in the Console, you'll get an error.** All Git
commands go in the Terminal.

```{r wrong-tab, eval=FALSE}
# THIS DOES NOT WORK — this is R, not the shell:
git status
#> Error: object 'git' not found

# Git commands go in the TERMINAL tab, not here!
```


# Verify Your Environment

Run this chunk to confirm R is working (this runs in the Console, which is
fine for R code):

```{r verify-r}
cat("R version:", R.version.string, "\n")
cat("Working directory:", getwd(), "\n")
cat("\n✓ R is working. Now let's check Git...\n")
```

Now switch to the **Terminal** tab and type these commands to verify Git:

```bash
# Type each of these in the Terminal (not here!)
git --version
git config --global --list
```

You should see a Git version number (2.x or higher). If `git config` shows
nothing, that's fine — we'll set it up in Module 2.


# About These Training Materials

These notebooks are designed to be used alongside the slide deck. The slides
provide the visual explanations; the notebooks provide the hands-on exercises
and reference material you can return to later.

All examples use NIEHS-relevant scenarios: PFAS biomonitoring, PM₂.₅ exposure
analysis, environmental health cohort studies, and dose-response modeling.

---

**Ready? Let's begin → open `01_why_version_control.Rmd`**
