---
title: "Module 1: Why Version Control for Science?"
subtitle: "The Problem, the Solution, and Why It Matters at NIEHS"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# The Problem: Sound Familiar?

How many of these have you seen in a project folder?

```
analysis.R
analysis_v2.R
analysis_v2_final.R
analysis_v2_final_REAL.R
analysis_v2_final_REAL_use_this.R
analysis_v2_final_REAL_use_this_FIXED.R
```

This is the "file name versioning" approach, and it fails in predictable ways:

- Which file is actually current?
- What changed between `v2` and `v2_final`?
- Who made the changes and why?
- Can you get back to last Tuesday's version?

**Version control eliminates this chaos entirely.**


# When Things Go Wrong Without Version Control

These are real scenarios (anonymized) from research settings:

### Scenario 1: The accidental deletion
A graduate student accidentally deleted 6 months of analysis code. No backup,
no history — they had to reconstruct from memory and old email attachments.

### Scenario 2: The overwrite
Two postdocs edited the same R script simultaneously on a shared drive. One
saved and overwrote the other's changes. Three days of work lost.

### Scenario 3: The reviewer question
A reviewer asked "how did you handle outliers in the 2nd revision?" The team
couldn't trace which version of the code produced which results.

### Scenario 4: The data leak
Sensitive patient identifiers were emailed in a data file. No audit trail of
who had access or when it was shared. Version control would show exactly when
the file was added and by whom.

> **Every one of these is preventable with Git.**


# What Git Gives You

| Benefit | What it means |
|---------|---------------|
| **Complete history** | Every change is recorded with who, what, when, and why — you can always go back |
| **Fearless experimentation** | Try new analyses on branches without risking your working code |
| **Reproducibility** | Collaborators and reviewers can see exactly how results were produced |
| **Collaboration** | Multiple scientists work on the same project without overwriting each other |


# Git vs. GitHub — They're Different Things

This is one of the most common points of confusion for beginners:

## Git

- The **version control software** that runs on your computer
- Tracks changes to files locally
- Works offline — no internet needed
- Free and open source, created in 2005 by Linus Torvalds (creator of Linux)
- On Posit Cloud, Git is pre-installed

## GitHub

- A **cloud hosting platform** for Git repositories
- Stores repos online for sharing
- Adds collaboration features: issues, pull requests, wikis, project boards
- Free for public and private repositories
- Owned by Microsoft since 2018
- Alternatives exist: GitLab, Bitbucket


# Comparing Approaches

How does Git compare to other ways of managing code?

```{r comparison-table, echo=FALSE}
comparison <- data.frame(
  Approach = c("Manual copies (analysis_v2_final.R)", 
               "Shared drive / Dropbox",
               "Email attachments",
               "Google Drive / OneDrive", 
               "Git + GitHub"),
  History = c("None", "Partial", "None", "Auto-save", "Full history"),
  Branching = c("None", "None", "None", "Limited", "Branches"),
  Collaboration = c("None", "Sync only", "Copies", "Comments", "PRs, Issues"),
  Verdict = c("Fragile", "Risky", "Chaotic", "No code tools", "Robust")
)

knitr::kable(comparison, align = "lcccc") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), 
                            full_width = FALSE)
```


# Where Git Fits in the Research Lifecycle

```
  Design     Collect     Write      Analyze &    Write        Publish &
  Study   →  Data     →  Code    →  Iterate   →  Manuscript → Archive
                         |__________________________________|
                                Git + GitHub active here
```

- Git tracks every change from the moment you start coding through publication
- The code that generated your published results should be traceable to a
  specific commit
- Many journals now require code and data availability statements — GitHub
  makes this easy


# Git Supports NIEHS FAIR+ Principles

The NIEHS 2025–2029 Strategic Plan emphasizes FAIR+ data and reproducibility.
Version control is foundational to both.

| Principle | How Git helps |
|-----------|---------------|
| **F**indable | Code in a public GitHub repository with DOI via Zenodo |
| **A**ccessible | Anyone can clone, inspect, and run your analysis |
| **I**nteroperable | Standard project structure works across platforms and tools |
| **R**eusable | Clear commit history documents every decision and change |


# Discussion Questions {.unnumbered}

Think about your own work:

1. Have you ever lost code or accidentally overwritten a colleague's changes?
2. If a reviewer asked you to show the exact code that produced Figure 3 in
   your last paper, could you find it?
3. How do you currently share code with collaborators?
4. What data security concerns do you have about putting code online?

We'll address all of these throughout the workshop.

---

**Next up → `02_setup.Rmd` — setting up GitHub, your PAT, and Posit Cloud**
