---
title: "Module 3: The Core Git Workflow"
subtitle: "Edit → Stage → Commit → Push"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# The Git Cycle

Every interaction with Git follows the same four-step cycle:

```
  Edit  →  Stage  →  Commit  →  Push
    |         |          |          |
  Make     Select     Record     Upload
  changes  changes    snapshot   to GitHub
  to files for save   locally    
```

You will repeat this cycle hundreds of times. It becomes muscle memory.


# The Three Areas of Git

Git manages your files across three areas. Understanding this model is the key
to understanding every Git command:

```
 Working Directory      Staging Area (Index)       Repository (History)
 ─────────────────      ────────────────────       ────────────────────
 Your actual files      Changes selected for       Permanent record of
 on disk — what         the next commit            all committed snapshots
 you edit

        ──── git add ────>       ──── git commit ────>
        <── git restore ──       <── git reset ──────
```

- **Working Directory:** The files you see and edit. Any file you modify shows
  up here as "changed."
- **Staging Area:** A holding zone where you select which changes go into the
  next commit. This is what makes Git powerful — you don't have to commit
  everything at once.
- **Repository:** The permanent history. Once committed, a snapshot is
  (essentially) permanent and can always be retrieved.


# `git status` — Your Most Important Command

Run `git status` constantly. It tells you exactly what state your files are in.

**Try it now in the Terminal:**

```bash
git status
```

The output tells you:

- What branch you're on
- Whether you're up to date with GitHub
- Which files are **staged** (green = ready to commit)
- Which files are **modified but not staged** (red = changed but not selected)
- Which files are **untracked** (red = new files Git doesn't know about yet)

### Reading `git status` output

```
On branch main
Your branch is up to date with 'origin/main'.

Changes to be committed:                    ← GREEN: staged
  modified:   README.md

Changes not staged for commit:              ← RED: changed but not staged
  modified:   analysis.R

Untracked files:                            ← RED: brand new files
  new_script.R
```


# Staging: Why Not Just Commit Everything?

Staging lets you craft **focused, meaningful commits** — even when you've
changed many files.

## Scenario

You've been working for an hour and changed 5 files:

1. Fixed a bug in the analysis script
2. Added a new figure
3. Updated the README
4. Cleaned up old commented-out code
5. Started (but didn't finish) a new model

### Without staging (bad)

```bash
git add .
git commit -m "updated a bunch of stuff"
# One giant commit with unrelated changes — impossible to review or revert
```

### With staging (good)

```bash
# Commit 1: the bug fix
git add analysis.R
git commit -m "Fix p-value calculation error in Table 2"

# Commit 2: the figure
git add R/03_visualize.R output/figures/fig3.png
git commit -m "Add Figure 3 dose-response curve"

# Commit 3: the README
git add README.md
git commit -m "Update methods description in README"

# DON'T commit the unfinished model yet — leave it for later
```

Three clean, reviewable, revertable commits instead of one messy blob.


# Stage & Commit: The RStudio GUI Method

1. **Check the Staged checkbox** in the Git pane next to files you want to
   include
2. Click **"Commit"** — opens the Review Changes window with a diff view
3. **Write a commit message** in the text box
4. Click the **"Commit"** button — Git records the snapshot
5. Click **"Push"** (↑ arrow) — uploads to GitHub

> When prompted for credentials, enter your GitHub **username** and your
> **PAT** (not your GitHub password).


# Stage & Commit: The Terminal Method

```bash
# Check what's changed
git status

# Stage specific files
git add analysis.R data/cleaned_data.csv

# Or stage everything (use with caution)
git add .

# Commit with a message
git commit -m "Add dose-response analysis for PFAS data"

# Push to GitHub
git push
```


# `git diff` — See Exactly What Changed

Before committing, review your changes line by line:

```bash
git diff                    # Show all unstaged changes
git diff analysis.R         # Show changes in one file
git diff --staged           # Show what's staged (about to be committed)
```

### Reading diff output

```
@@ -12,7 +12,8 @@ library(tidyverse)

 # Fit the model
-model <- lm(outcome ~ dose, data = df)
+model <- lm(outcome ~ dose + age + sex, data = df)
+# Adjusted for key confounders per reviewer feedback
 summary(model)
```

- **Red (`-`)** = removed line
- **Green (`+`)** = added line
- **White** = unchanged context

You can also see diffs in the RStudio GUI: click **"Diff"** in the Git pane,
or click a filename to open the diff viewer.


# Writing Good Commit Messages

Your commit messages are documentation for your future self and your
collaborators. They answer: **what changed and why?**

## Bad commit messages

```
"update"
"fix stuff"  
"asdfghjkl"
"changes"
"final version"
"misc"
```

## Good commit messages

```
"Add PFAS dose-response model"
"Fix p-value calculation in Figure 3"
"Remove hardcoded file paths"
"Add methods section to README"
"Filter outliers in PM2.5 data"
"Adjust for BMI confounding"
```

## The Rule

Complete the sentence: **"This commit will..."**

- "This commit will **add PFAS dose-response model**" ✓
- "This commit will **update**" ✗ (update what? why?)

## Anatomy of a Great Commit Message

```
Adjust dose-response model for age and BMI confounders

Reviewer 2 requested adjustment for potential confounders
in the PFAS-cholesterol analysis. Added age and BMI to
the linear model in analysis/03_regression.R.

Changes:
- Added age (continuous) and BMI (continuous) as covariates
- Updated Table 2 to show adjusted estimates
- R-squared improved from 0.12 to 0.31

Closes #7
```

- **First line:** Subject — imperative mood, under 50 characters
- **Blank line** separates subject from body
- **Body:** Explains why, not just what
- **"Closes #7":** Automatically closes GitHub issue #7 when merged


# Viewing History: `git log`

```bash
git log --oneline              # Compact one-line-per-commit view
git log --oneline -5           # Last 5 commits only
git log --stat                 # Show which files changed per commit
git log --author="Smith"       # Commits by a specific person
git log -- analysis.R          # Commits touching a specific file
```

### Example output

```
a3f2d1c Adjust model for age and BMI confounders
b7e4a92 Add Figure 3 dose-response curve
c1d8f3e Fix p-value calculation error
d9a2b7f Import cleaned PFAS biomonitoring data
e5c1d4a Initial commit with project structure
```

Each commit has a unique **hash** (like `a3f2d1c`) that you can use to
reference it.

> **In the RStudio GUI:** Click the clock icon (⏱) in the Git pane to see the
> visual commit history.


# Undoing Mistakes

Git's safety net is one of its biggest advantages. Here are the most common
"undo" operations:

## Unstage a file (remove from staging, keep changes)

```bash
git restore --staged filename.R
```

## Discard changes to a file (revert to last commit)

```bash
git restore filename.R
# ⚠️ DESTRUCTIVE — your changes are gone
```

## Undo the last commit (keep the changes staged)

```bash
git reset --soft HEAD~1
```

## View an old version of a file

```bash
git show HEAD~2:analysis.R    # 2 commits ago
git show a3f2d1c:analysis.R   # Specific commit by hash
```

## Fix the last commit message

```bash
git commit --amend
# Opens an editor to change the message (only works before pushing)
```


# `git stash` — Save Work Temporarily

Need to switch branches but have uncommitted changes? **Stash** saves your
work-in-progress without committing.

```bash
# Save current uncommitted work
git stash push -m "WIP: new exposure model"

# Switch branches, do other work...
git checkout main

# Come back and restore your work
git checkout feature-branch
git stash pop                  # Applies and removes from stash
```

Other stash commands:

```bash
git stash list                 # View all stashes
git stash drop                 # Delete the most recent stash
git stash apply                # Apply without removing from stash
```


# .gitignore Preview

Before committing files, set up `.gitignore` to exclude things that don't
belong in version control:

```
# R workspace & history
.RData
.Rhistory
.Rproj.user/

# Large data files
data/raw/*.csv
*.rds

# Sensitive info
.Renviron
credentials/

# Regeneratable output
output/figures/*.png
```

When you created the repo with GitHub's R template, many of these were already
added for you. We'll cover `.gitignore` in detail in Module 6.


# Exercise 1: Your First Commit Cycle {.unnumbered}

**Time: 10 minutes**

1. In Posit Cloud, create a new R script: **File → New File → R Script**
2. Write 3–5 lines of R code:

```{r exercise1-example, eval=FALSE}
# My first tracked script
library(datasets)
data(iris)
summary(iris$Sepal.Length)
hist(iris$Sepal.Length, main = "Sepal Length Distribution",
     col = "steelblue", xlab = "Length (cm)")
```

3. Save it as `my_first_script.R`
4. In the **Terminal**: `git status` — see it listed as untracked
5. In the **Git pane**: check the **Staged** box next to `my_first_script.R`
6. Click **Commit** → write message: `"Add first practice script"` → **Commit**
7. Click **Push** (↑) → enter username + PAT when prompted
8. Go to your GitHub repo in a browser — confirm the file appeared!


# Exercise 2: Modify, Diff, and Commit Again {.unnumbered}

**Time: 8 minutes**

1. Open `my_first_script.R` and **add 3 more lines** of code:

```{r exercise2-additions, eval=FALSE}
# Added: summary statistics
cat("Mean sepal length:", mean(iris$Sepal.Length), "\n")
cat("SD:", sd(iris$Sepal.Length), "\n")
cat("Number of species:", length(unique(iris$Species)), "\n")
```

2. Save the file
3. In the Terminal: `git diff my_first_script.R` — see the green `+` lines
4. In the Git pane: click the filename to see the diff view
5. Stage the file (check the box)
6. Commit with message: `"Expand script with summary statistics"`
7. Push to GitHub
8. On GitHub: click the file → click **"History"** → see both commits
9. **Bonus:** In the Terminal, try `git log --oneline`

---

**Time for a break! Next up → `04_branching.Rmd`**
