---
title: "Module 4: Branching & Collaboration"
subtitle: "Parallel Experiments, Pull Requests, and Merge Conflicts"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What Are Branches?

A branch is a **separate line of development**. Think of it as a lab notebook
page you can tear out if the experiment fails — without affecting your main
results.

```
main:       ●───●───●───●───●───●
                 \              ↑
                  \    merge ──┘
new-model:         ●───●───●
```

- **`main`** is your stable, working code
- You create a **feature branch** to try something new
- If it works, you **merge** it back into main
- If it doesn't work, you just delete the branch — main is untouched


# Branch Commands

```bash
# Create a new branch AND switch to it
git checkout -b new-model

# List all branches (* = current)
git branch

# Switch between existing branches
git checkout main
git checkout new-model

# Merge a branch into your current branch
# (first switch to the branch you want to merge INTO)
git checkout main
git merge new-model

# Delete a branch (after it's merged)
git branch -d new-model
```


# Branch Naming Conventions

Good branch names describe what you're working on. Keep them short, lowercase,
with hyphens.

| Pattern | Example | Use for |
|---------|---------|---------|
| `feature/...` | `feature/add-sensitivity-analysis` | New functionality |
| `fix/...` | `fix/p-value-calculation` | Bug fixes |
| `docs/...` | `docs/update-readme-methods` | Documentation |
| `explore/...` | `explore/random-forest-model` | Experiments (may discard) |
| `review/...` | `review/reviewer2-confounders` | Revision responses |

**Avoid:** spaces, uppercase, vague names like `test`, `my-branch`, or `temp`.


# Working on a Branch — Full Workflow

Here is the complete workflow from creating a branch to merging it:

```bash
# 1. Create and switch to a new branch
git checkout -b feature/add-sensitivity-analysis

# 2. Do your work — edit files, write code, analyze
#    (work normally; Git tracks changes on this branch)

# 3. Stage and commit (possibly multiple times)
git add analysis/sensitivity.R
git commit -m "Add sensitivity analysis for dose threshold"

git add output/figures/fig_sensitivity.png
git commit -m "Add sensitivity analysis figure"

# 4. Push the branch to GitHub
git push -u origin feature/add-sensitivity-analysis
# The -u flag sets up tracking so future pushes just need 'git push'

# 5. Open a Pull Request on GitHub (in the browser)

# 6. After review & merge, switch back and pull
git checkout main
git pull
```


# Staying In Sync: Pull Before You Push

When collaborating, always pull the latest changes before starting new work.

```
Your Daily Workflow
───────────────────
1. git pull           ← Get latest from GitHub
2. (do your work)     ← Edit, analyze, write
3. git add .          ← Stage changes
4. git commit -m "…"  ← Record snapshot
5. git push           ← Share with team
```

> **Forgetting to pull** is the most common cause of merge conflicts. Make it a
> habit: always pull first.


# Pull Requests (PRs)

A Pull Request is a **request to merge your branch into main**. It's where
collaboration happens on GitHub.

## The PR Workflow

1. **Push your branch** to GitHub: `git push -u origin branch-name`
2. **Open a PR on GitHub** — click the "Compare & pull request" banner
3. **Describe your changes** — what, why, and what reviewers should check
4. **Request review** — tag a collaborator
5. **Discuss & iterate** — reviewer leaves comments, you make fixes and push
6. **Merge!** — click "Merge pull request" when approved

## Writing a Good PR Description

```markdown
## What this PR does
Adds sensitivity analysis varying the PFAS dose-response
threshold from 2.0 to 8.0 ng/mL in 0.5 increments.

## Why
Reviewer 2 requested this to assess robustness of our
primary finding. See Issue #7.

## What to review
- analysis/sensitivity.R — new script, ~80 lines
- output/figures/fig_sensitivity.png — new figure
- Does the threshold range seem appropriate?

## Testing
Knitted full report successfully. All existing tests pass.
```


# Merge Conflicts

Conflicts happen when two people (or two branches) edit the **same line** of
the same file. Git can't decide which version to keep, so it asks you.

## What a conflict looks like

```
<<<<<<< HEAD
model <- lm(outcome ~ dose + age, data = df)
=======
model <- lm(outcome ~ dose * age, data = df)
>>>>>>> new-model

# YOUR version (HEAD) is above =======
# THEIR version (new-model) is below =======
```

## Resolving step by step

1. **Git tells you there's a conflict** — "CONFLICT" in the terminal output
2. **Open the conflicted file** — look for `<<<<<<<` and `>>>>>>>` markers
3. **Decide which version to keep** — yours, theirs, or a combination
4. **Delete ALL conflict markers** — remove `<<<<<<<`, `=======`, `>>>>>>>`
5. **Test that the file works** — run the code!
6. **Stage and commit the resolution:**

```bash
git add analysis.R
git commit -m "Resolve merge conflict in model specification"
```

## Prevention

- **Communicate** with your team about who's editing which files
- **Pull often** — the longer you wait, the more divergence
- **Use branches** — they isolate changes and make conflicts easier to handle
- **Keep commits small** — smaller changes = smaller conflicts


# Cleaning Up: Deleting Merged Branches

After a branch is merged, delete it to keep your repo tidy. The history is
preserved in main — the branch is just a label.

```bash
# Delete a local branch (after it's merged)
git branch -d feature/add-sensitivity

# Delete the remote branch on GitHub
git push origin --delete feature/add-sensitivity

# List all branches (local and remote)
git branch -a
```

> GitHub also offers a **"Delete branch"** button on the PR page after
> merging — this is the easiest way.


# Exercise 3: Branch, Pull Request, Merge {.unnumbered}

**Time: 10 minutes**

1. **Create a branch:**
```bash
git checkout -b add-exploration
```

2. **Create a new R script** (`exploration.R`):

```{r exercise3-code, eval=FALSE}
# Exploring the mtcars dataset
library(ggplot2)
data(mtcars)

# Quick summary
summary(mtcars[, c("mpg", "hp", "wt")])

# Scatter plot
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(title = "Fuel Efficiency by Weight",
       x = "Weight (1000 lbs)", y = "MPG",
       color = "Cylinders") +
  theme_minimal()
```

3. **Stage and commit:**
```bash
git add exploration.R
git commit -m "Add mtcars exploration with scatter plot"
```

4. **Push the branch:**
```bash
git push -u origin add-exploration
```

5. **On GitHub.com:**
   - Click **"Compare & pull request"**
   - Write a short description
   - Click **"Create pull request"**
   - Click **"Merge pull request"** → **"Confirm merge"**
   - Click **"Delete branch"**

6. **Back in Posit Cloud Terminal:**
```bash
git checkout main
git pull
git branch -d add-exploration
```

7. Verify: your `exploration.R` file should now be on main!

---

**Next up → `05_github_for_teams.Rmd`**
