---
title: "Module 5: GitHub for Scientific Teams"
subtitle: "Issues, READMEs, Releases, DOIs, Pages, and Licensing"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# GitHub Issues: Your Research To-Do List

Issues track bugs, tasks, ideas, and discussion — all linked to your code.

## What makes Issues powerful

- **Linked to code:** Reference issues in commits and PRs
- **Assignable:** Assign tasks to specific team members
- **Labeled:** Categorize by type (bug, enhancement, question)
- **Searchable:** Full-text search across all issues
- **Closeable:** Mark as done when resolved

## Creating an Issue

On GitHub: **Repo → Issues tab → New issue**

A good issue includes:

- **Clear title:** "Figure 2 colors not colorblind-friendly"
- **Description:** What's wrong, what's expected, how to reproduce
- **Labels:** bug, enhancement, documentation, etc.
- **Assignee:** Who should work on this?

## Closing Issues with Commits

Use keywords in commit messages to auto-close issues:

```bash
git commit -m "Fix colorblind palette for Figure 2, closes #11"
```

Keywords that work: `closes`, `fixes`, `resolves` (plus their variants:
`close`, `fix`, `resolve`, `closed`, `fixed`, `resolved`)

The issue closes automatically when the commit is merged to main.

## Useful Labels

| Label | Description | Color suggestion |
|-------|-------------|-----------------|
| `bug` | Something isn't working | Red |
| `enhancement` | New feature or analysis | Blue |
| `documentation` | Docs, README, methods | Green |
| `question` | Discussion or decision needed | Purple |
| `good first issue` | Simple task for new contributors | Violet |
| `help wanted` | Extra attention needed | Teal |
| `priority: high` | Needs immediate attention | Dark red |


# README.md: Your Project's Front Page

GitHub renders `README.md` as the landing page. A good README answers:
**what is this project, how do I use it, and who made it?**

## Template

````markdown
# PFAS Dose-Response Analysis

Analysis of PFAS biomonitoring data from the RTP cohort.

## Overview

This project examines dose-response relationships between
serum PFAS concentrations and metabolic outcomes in 482
participants from the Research Triangle Park area.

## Getting Started

1. Clone the repo: `git clone https://github.com/...`
2. Install dependencies: `renv::restore()`
3. Run the full analysis: `source("R/00_run_all.R")`

## Project Structure

```
R/              Analysis scripts (numbered in order)
data/raw/       Original data files (not tracked in Git)
data/processed/ Cleaned datasets
output/figures/ Generated figures
output/tables/  Generated tables
```

## Authors

- Jane Smith, Ph.D. — Biostatistics Branch, NIEHS
- John Doe, Ph.D. — Epidemiology Branch, NIEHS

## License

MIT License — see LICENSE file for details.
````


# Organizing a Scientific Repository

A consistent project structure makes your work navigable for collaborators
and reviewers.

```
my-project/
├── R/
│   ├── 01_clean.R          # Data cleaning
│   ├── 02_analyze.R        # Statistical analysis
│   └── 03_visualize.R      # Figures and tables
├── data/
│   ├── raw/                # Original data (often gitignored)
│   └── processed/          # Cleaned data
├── output/
│   ├── figures/            # Generated plots
│   └── tables/             # Generated tables
├── docs/                   # Documentation, manuscripts
├── .gitignore              # Files to exclude
├── README.md               # Project description
├── LICENSE                 # How others can use your code
└── my-project.Rproj        # RStudio project file
```

**Key principles:**

- **Number your scripts** — `01_`, `02_` prefixes make the analysis order clear
- **Separate raw from processed** — never modify raw data files
- **Use an R Project (.Rproj)** — keeps paths relative and portable
- **Include a README** — future you (and reviewers) will thank you


# GitHub Projects: Visual Task Management

GitHub Projects provides a **Kanban board** view of your issues — like Trello,
but integrated with your code.

```
┌──────────┐  ┌──────────┐  ┌──────────┐  ┌──────────┐
│  To Do   │  │In Progress│  │In Review │  │   Done   │
├──────────┤  ├──────────┤  ├──────────┤  ├──────────┤
│ Issue #5 │  │ Issue #3 │  │  PR #8   │  │ Issue #1 │
│ Issue #6 │  │ Issue #7 │  │          │  │ Issue #2 │
│ Issue #9 │  │          │  │          │  │ Issue #4 │
└──────────┘  └──────────┘  └──────────┘  └──────────┘
```

**Setup:** Repo → Projects tab → New project → Board view

Issues can be dragged between columns. PRs linked to issues auto-move to
"Done" when merged.


# Releases & Tags: Marking Milestones

Tags mark specific commits. Releases bundle a tag with notes and downloads.

## Creating a tag and release

```bash
# Create an annotated tag
git tag -a v1.0 -m "Version submitted to EHP"
git push origin v1.0
```

Then on GitHub: **Repo → Releases → Draft a new release → Choose tag `v1.0`**

## When to create releases

| Milestone | Tag | Notes |
|-----------|-----|-------|
| Manuscript submission | `v1.0` | Exact code that produced submitted results |
| Manuscript revision | `v2.0` | Updated analysis per reviewer comments |
| Data freeze | `v0.9-data-freeze` | Point when data cleaning was finalized |
| Conference presentation | `v1.0-asc2025` | Code behind your poster or talk |


# Zenodo + GitHub = Citable Code

[Zenodo](https://zenodo.org) (by CERN) archives your GitHub releases and
assigns a **DOI** — making your code citable in publications.

## Setup (one-time, 5 minutes)

1. Go to [zenodo.org](https://zenodo.org) and log in with your GitHub account
2. Go to Settings → GitHub → flip the toggle **ON** for your repository
3. Create a **Release** on GitHub (as described above)
4. Zenodo automatically archives it and assigns a DOI
5. Add the DOI badge to your README:

```markdown
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.XXXXXXX.svg)](https://doi.org/10.5281/zenodo.XXXXXXX)
```

> **Many journals now require a DOI for code and data.** Zenodo + GitHub makes
> this a 5-minute task, not a bureaucratic ordeal.


# Forking: Contributing to Others' Projects

A **fork** is your personal copy of someone else's repository. You can modify
it freely without affecting the original.

## Fork workflow

1. Click **"Fork"** on someone's GitHub repository
2. **Clone YOUR fork** to Posit Cloud
3. Create a branch and make changes
4. Push to YOUR fork
5. Open a **Pull Request from your fork** to the original repo
6. The maintainer reviews and merges (or requests changes)

**Useful for:** contributing to open-source R packages, building on a
collaborator's published analysis, adapting someone's methods for your data.


# GitHub Pages: Free Project Websites

GitHub Pages turns a repository into a website — perfect for project
documentation, supplementary materials, and lab pages.

## Setup

1. Repo → Settings → Pages → Source: "Deploy from a branch" → main → `/docs`
2. Add HTML or Markdown files to the `/docs` folder
3. Commit and push — your site goes live at `username.github.io/repo-name`

**R Markdown tip:** Knit your report to HTML, save in `/docs`, push to GitHub,
and you have an instant supplementary methods page that anyone can view.


# Licensing Your Code

Without a license, your code is **"all rights reserved"** by default — others
legally cannot reuse it, even if it's on a public GitHub repo.

```{r license-table, echo=FALSE}
licenses <- data.frame(
  License = c("MIT License", "GPL-3.0", "CC-BY-4.0", "No license"),
  Meaning = c("Do anything, just keep the copyright notice",
              "Derivatives must also be open source",
              "For non-code content (docs, data); requires attribution",
              "All rights reserved — no reuse allowed"),
  Note = c("Most permissive, widely used",
           "'Copyleft' — ensures code stays open",
           "Good for datasets and documentation",
           "Avoid this for reproducible science!")
)
knitr::kable(licenses) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            full_width = FALSE)
```

**To add:** Create a `LICENSE` file in your repo, or select one when creating
the repo on GitHub.

> **Check with your NIEHS tech transfer office** about policies for
> government-funded code. Many NIH-funded projects use public domain or MIT
> licensing.

---

**Next up → `06_best_practices.Rmd`**
