---
title: "Module 6: Best Practices & .gitignore"
subtitle: "Security, Commit Hygiene, and Keeping Your Repo Clean"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# .gitignore: What NOT to Track

A `.gitignore` file tells Git which files to **automatically exclude** from
version control. It lives in the root of your repository.

## Essential .gitignore for R Projects

```
# R workspace and history — these are user-specific
.RData
.Rhistory
.Rproj.user/

# Large data files — too big for Git (>50 MB)
data/raw/*.csv
data/raw/*.xlsx
*.rds
*.rda

# Sensitive information — NEVER commit these
.Renviron
credentials/
*.pem
*.key

# Output files that can be regenerated
output/figures/*.png
output/figures/*.tiff
output/tables/*.csv

# OS-specific junk
.DS_Store
Thumbs.db

# Temporary files
*~
*.swp
```

> When you created the repo with GitHub's **R .gitignore template**, many of
> these patterns were already included. Always review and customize.


# .gitignore Pattern Syntax

```{r pattern-table, echo=FALSE}
patterns <- data.frame(
  Pattern = c("*.csv", "data/*.csv", "data/**/*.csv", "!data/reference.csv",
              "output/", ".Renviron", "*.log", "temp_*"),
  Effect = c("Ignore all CSV files everywhere",
             "Ignore CSVs in the data/ folder only",
             "Ignore CSVs in data/ and all subfolders",
             "Exception: DO track this specific CSV",
             "Ignore the entire output directory",
             "Ignore this specific file",
             "Ignore all log files",
             "Ignore files starting with temp_")
)
knitr::kable(patterns) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            full_width = FALSE)
```

### Key rules

- Patterns are matched from the repo root
- `*` matches any characters within a filename
- `**` matches across directories
- `!` creates an **exception** (override a previous ignore pattern)
- Order matters — later rules override earlier ones
- Lines starting with `#` are comments
- Directories end with `/`


## What to Track vs. What to Ignore

**DO track:**

- R scripts (`.R`), R Markdown (`.Rmd`), Quarto (`.qmd`)
- README.md and documentation
- `.gitignore` itself
- Small reference datasets (if not sensitive)
- Configuration files (`.Rproj`, `renv.lock`)
- LICENSE file

**DON'T track:**

- Large datasets (>50 MB) — use data repositories instead
- Passwords, API keys, tokens — use `.Renviron`
- `.RData` and `.Rhistory` — user-specific, often huge
- Generated output you can regenerate from code
- Protected health information (PHI) or personally identifiable data (PII)


## The .Renviron Pattern for Secrets

Store API keys and passwords in `.Renviron` (which is gitignored):

```
# .Renviron file (in your project root)
API_KEY=your_secret_key_here
DATABASE_PASSWORD=another_secret
```

Access in your R code:

```{r renviron-demo, eval=FALSE}
# In your R script
my_key <- Sys.getenv("API_KEY")
```

This keeps secrets out of your code AND out of version control.


# Security for Federal Scientists

<div style="padding: 12px; background-color: #FEE2E2; border-left: 4px solid #DC2626; margin: 16px 0;">
**Critical:** Once data is pushed to GitHub, it is in the repository history —
**even if you delete the file later.** Removing a file from Git history requires
rewriting history, which is complex and disruptive.
</div>

## Security Rules

1. **Never commit sensitive data** — No PII, PHI, API keys, or unpublished
   confidential data
2. **Use private repositories** — Free accounts support private repos. Make
   repos public only when ready for publication
3. **Use .gitignore proactively** — Add patterns BEFORE creating sensitive
   files. Prevention is far easier than cleanup
4. **Use .Renviron for secrets** — Store API keys there (gitignored) and
   access via `Sys.getenv()` in your scripts
5. **Check with your ISSO** — Your NIEHS Information System Security Officer
   can advise on institute-specific policies
6. **Audit before making public** — Review ALL files and the full commit
   history before changing repository visibility from private to public

## What if sensitive data was accidentally committed?

1. **Don't panic**, but act quickly
2. If the repo is private and hasn't been shared, you may be able to rewrite
   history with `git filter-branch` or the BFG Repo Cleaner
3. If the repo is public, **assume the data is compromised** — rotate any
   exposed credentials immediately
4. Contact your ISSO and follow NIEHS incident response procedures
5. Consider deleting the repository and recreating it


# How Often Should You Commit?

Think of commits as saving a document — but with purpose. Each commit should
be a **logical unit of work**.

### Too infrequent (bad)

- One commit per week
- Hundreds of changes mixed together
- Impossible to review or revert individual changes
- Defeats the purpose of Git

### Just right (good)

- After each logical change
- 2–10 commits per work session
- Each commit has a clear, describable purpose
- Can be reviewed independently

**Rule of thumb:** If you can't describe the commit in one sentence, it's
probably too big. When in doubt, commit more often.


# Other Best Practices

## Use relative paths, not absolute paths

```{r paths-demo, eval=FALSE}
# BAD — breaks on anyone else's computer
data <- read.csv("/Users/jsmith/Desktop/niehs/data/exposure.csv")

# GOOD — works from the project root on any machine
data <- read.csv("data/processed/exposure.csv")

# BETTER — works even if your working directory changes
data <- read.csv(here::here("data", "processed", "exposure.csv"))
```

## Use R Projects (.Rproj)

Always work within an R Project. This:

- Sets the working directory to the project root
- Makes relative paths work consistently
- Enables Git integration in the RStudio GUI
- Keeps your R environment separate from other projects

## Use `renv` for package management

```{r renv-demo, eval=FALSE}
# Initialize renv in your project
renv::init()

# After installing packages, take a snapshot
renv::snapshot()

# A collaborator restores your exact package versions
renv::restore()
```

The `renv.lock` file records every package and version. Commit it to Git!
This ensures anyone can reproduce your environment.

## Write code for humans, not just computers

```{r readable-code, eval=FALSE}
# BAD
m<-lm(y~x1+x2+x3,d);s<-summary(m);s$r.squared

# GOOD
model <- lm(
  cholesterol ~ pfoa_serum + age + bmi,
  data = biomarker_data
)
model_summary <- summary(model)
model_summary$r.squared
```


# Quick Reference: Git Commands {.tabset}

## Setup

```bash
git config --global user.name "Your Name"
git config --global user.email "you@niehs.nih.gov"
git config --global --list
```

## Daily Workflow

```bash
git status                # What changed?
git add <file>            # Stage a file
git add .                 # Stage everything
git commit -m "message"   # Commit with message
git push                  # Upload to GitHub
git pull                  # Download from GitHub
```

## Branching

```bash
git checkout -b <name>    # Create & switch to branch
git checkout main         # Switch to main
git branch                # List branches
git merge <branch>        # Merge branch into current
git branch -d <branch>    # Delete merged branch
```

## Inspection

```bash
git log --oneline         # Compact history
git log --oneline -5      # Last 5 commits
git diff                  # Unstaged changes
git diff --staged         # Staged changes
git show HEAD:file.R      # View old file version
```

## Undoing

```bash
git restore --staged f.R  # Unstage a file
git restore f.R           # Discard changes (destructive!)
git reset --soft HEAD~1   # Undo last commit (keep changes)
git stash                 # Save work temporarily
git stash pop             # Restore stashed work
git commit --amend        # Fix last commit message
```

## {.unnumbered}

---

**Next up → `07_capstone.Rmd` — put it all together!**
