---
title: "Capstone Exercise: Full Git Workflow"
subtitle: "Put Everything Together in 15 Minutes"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview {.unnumbered}

This capstone exercise walks you through a complete Git workflow that uses
every skill from the workshop. You will:

1. Create a branch
2. Write an analysis script
3. Stage and commit with a meaningful message
4. Push the branch to GitHub
5. Open and merge a Pull Request
6. Create and close a GitHub Issue
7. Sync your local repo

**Time: 15 minutes.** Work at your own pace. The steps below are detailed —
follow them in order.

---

# Step 1: Create a Branch

In the **Terminal:**

```bash
git checkout -b feature/capstone-analysis
```

Verify you're on the new branch:

```bash
git branch
# Should show * feature/capstone-analysis
```


# Step 2: Write an Analysis Script

Create a new file called `capstone_analysis.R` with the following content.
You can type it manually or copy from this notebook:

```{r capstone-code, eval=FALSE}
# ==============================================================
# Capstone Analysis: Environmental Exposure Summary
# Author: [YOUR NAME]
# Date: [TODAY'S DATE]
# Description: Quick exploration of simulated exposure data
# ==============================================================

library(ggplot2)

# --- Simulated biomonitoring data ---
set.seed(42)
n <- 100
exposure_data <- data.frame(
  participant_id = paste0("NIEHS-", sprintf("%04d", 1:n)),
  age = round(rnorm(n, mean = 48, sd = 13)),
  sex = sample(c("Female", "Male"), n, replace = TRUE),
  pfoa_ng_ml = round(rlnorm(n, meanlog = 0.8, sdlog = 0.9), 2),
  site = sample(c("RTP", "Durham", "Chapel Hill"), n, replace = TRUE)
)

# --- Summary statistics ---
cat("=== PFOA Serum Concentrations ===\n")
cat("N:", nrow(exposure_data), "\n")
cat("Mean:", round(mean(exposure_data$pfoa_ng_ml), 2), "ng/mL\n")
cat("Median:", round(median(exposure_data$pfoa_ng_ml), 2), "ng/mL\n")
cat("Range:", round(range(exposure_data$pfoa_ng_ml), 2), "ng/mL\n")

# --- Visualization ---
p <- ggplot(exposure_data, aes(x = pfoa_ng_ml, fill = sex)) +
  geom_histogram(alpha = 0.7, bins = 25, position = "identity") +
  scale_fill_manual(values = c("Female" = "#0D9488", "Male" = "#6366F1")) +
  scale_x_log10() +
  labs(
    title = "Distribution of Serum PFOA Concentrations",
    subtitle = paste("N =", n, "participants"),
    x = "PFOA (ng/mL, log scale)",
    y = "Count",
    fill = "Sex"
  ) +
  theme_minimal()

print(p)

cat("\nAnalysis complete.\n")
```

Save the file as `capstone_analysis.R`.


# Step 3: Check Status and Diff

In the Terminal:

```bash
# See the new untracked file
git status

# If you also modified any existing files:
git diff
```


# Step 4: Stage and Commit

```bash
# Stage the new file
git add capstone_analysis.R

# Commit with a descriptive message
git commit -m "Add capstone analysis of simulated PFOA biomonitoring data"
```


# Step 5: Push the Branch to GitHub

```bash
git push -u origin feature/capstone-analysis
```

Enter your username and PAT when prompted.


# Step 6: Open a Pull Request on GitHub

1. Go to your repository on GitHub.com
2. You should see a yellow banner: **"feature/capstone-analysis had recent
   pushes"** → click **"Compare & pull request"**
3. Fill in the PR description:

```
## What this PR does
Adds a capstone analysis script that explores simulated
PFOA biomonitoring data with summary statistics and a
histogram stratified by sex.

## Skills demonstrated
- Branching
- Staging and committing
- Writing descriptive commit messages
- Pull request workflow
```

4. Click **"Create pull request"**


# Step 7: Merge the Pull Request

1. On the PR page, click **"Merge pull request"**
2. Click **"Confirm merge"**
3. Click **"Delete branch"** (cleanup)


# Step 8: Sync Your Local Repo

Back in the Posit Cloud Terminal:

```bash
git checkout main
git pull
```

Your `capstone_analysis.R` should now be on main. Verify:

```bash
ls -la capstone_analysis.R
git log --oneline -3
```


# Step 9: Create and Close an Issue

1. On GitHub: **Repo → Issues → New issue**
2. Title: `"Add boxplot of PFOA by site"`
3. Description: `"The capstone analysis should include a boxplot comparing PFOA concentrations across the three monitoring sites."`
4. Click **"Submit new issue"** — note the issue number (probably `#1`)

Now close it with a commit:

```bash
# Make sure you're on main
git checkout main

# Add a line to capstone_analysis.R:
echo '
# Boxplot by site
p2 <- ggplot(exposure_data, aes(x = site, y = pfoa_ng_ml, fill = site)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_log10() +
  labs(title = "PFOA by Monitoring Site", y = "PFOA (ng/mL, log scale)") +
  theme_minimal() +
  theme(legend.position = "none")
print(p2)
' >> capstone_analysis.R

# Stage and commit — closes the issue!
git add capstone_analysis.R
git commit -m "Add PFOA boxplot by site, closes #1"
git push
```

Go to GitHub → Issues → verify that issue #1 is now closed!


# Congratulations! {.unnumbered}

<div style="padding: 15px; background-color: #F0FDF4; border-left: 4px solid #0D9488; margin-top: 20px;">

If you completed this capstone, you have practiced:

- ✅ Creating and switching branches
- ✅ Writing and saving R scripts
- ✅ Staging and committing with meaningful messages
- ✅ Pushing branches to GitHub
- ✅ Opening and merging Pull Requests
- ✅ Creating GitHub Issues
- ✅ Closing Issues from commit messages
- ✅ Syncing your local repo with `git pull`
- ✅ Deleting merged branches

**You are now equipped to use Git and GitHub in your daily research work.**

The best way to learn Git is to **use it every day** — start with your next
project, even if it's solo work.

</div>
